---
title: "STAT 184 Final Project" 
subtitle: "EDA"
author: "Jonathan Sguera"
output: html_notebook
---



## Guiding Question: 
Do English Premier League matches have higher Full Time Total goals on average than Italian Serie A matches?

```{r}
# clean up workspace environment
rm(list = ls())

# all packages used for the assignment
library(mosaic)
library(tidyverse)
library(DataComputing)

```

## Acquainted with Data sources:
```{r}
# Here I am reading in the data from the website link I found it at
Serie_A_Data <- read.csv("https://datahub.io/sports-data/italian-serie-a/r/season-1819.csv")
Premier_League_Data <- read.csv("https://datahub.io/sports-data/english-premier-league/r/season-1819.csv")
```

### Where did you find them?
I found these data tables from datahub.io and I only included data from the 2018-19 season.

### Who collected/maintains them?
Data comes from football-data.co.uk which is a English website which gathers match data including betting odds to help gamblers decide which bets to make. 
### When & why were they originally collected?
* Below I have added a key from football-data.co.uk which explains where all data came from and includes betting odds from multiple websites. THe original data seems to have been collected to help fans determine which matches to bet on and which website to use for avid European Soccer gamblers.

Key to results data:

Div = League Division
Date = Match Date (dd/mm/yy)
HomeTeam = Home Team
AwayTeam = Away Team
FTHG = Full Time Home Team Goals
FTAG = Full Time Away Team Goals
FTR = Full Time Result (H=Home Win, D=Draw, A=Away Win)
HTHG = Half Time Home Team Goals
HTAG = Half Time Away Team Goals
HTR = Half Time Result (H=Home Win, D=Draw, A=Away Win)

Match Statistics (where available)
Attendance = Crowd Attendance
Referee = Match Referee
HS = Home Team Shots
AS = Away Team Shots
HST = Home Team Shots on Target
AST = Away Team Shots on Target
HHW = Home Team Hit Woodwork
AHW = Away Team Hit Woodwork
HC = Home Team Corners
AC = Away Team Corners
HF = Home Team Fouls Committed
AF = Away Team Fouls Committed
HO = Home Team Offsides
AO = Away Team Offsides
HY = Home Team Yellow Cards
AY = Away Team Yellow Cards
HR = Home Team Red Cards
AR = Away Team Red Cards
HBP = Home Team Bookings Points (10 = yellow, 25 = red)
ABP = Away Team Bookings Points (10 = yellow, 25 = red)

Key to 1X2 (match) betting odds data:

B365H = Bet365 home win odds
B365D = Bet365 draw odds
B365A = Bet365 away win odds
BSH = Blue Square home win odds
BSD = Blue Square draw odds
BSA = Blue Square away win odds
BWH = Bet&Win home win odds
BWD = Bet&Win draw odds
BWA = Bet&Win away win odds
GBH = Gamebookers home win odds
GBD = Gamebookers draw odds
GBA = Gamebookers away win odds
IWH = Interwetten home win odds
IWD = Interwetten draw odds
IWA = Interwetten away win odds
LBH = Ladbrokes home win odds
LBD = Ladbrokes draw odds
LBA = Ladbrokes away win odds
PSH = Pinnacle Sports home win odds
PSD = Pinnacle Sports draw odds
PSA = Pinnacle Sports away win odds
SOH = Sporting Odds home win odds
SOD = Sporting Odds draw odds
SOA = Sporting Odds away win odds
SBH = Sportingbet home win odds
SBD = Sportingbet draw odds
SBA = Sportingbet away win odds
SJH = Stan James home win odds
SJD = Stan James draw odds
SJA = Stan James away win odds
SYH = Stanleybet home win odds
SYD = Stanleybet draw odds
SYA = Stanleybet away win odds
VCH = VC Bet home win odds
VCD = VC Bet draw odds
VCA = VC Bet away win odds
WHH = William Hill home win odds
WHD = William Hill draw odds
WHA = William Hill away win odds

Bb1X2 = Number of BetBrain bookmakers used to calculate match odds averages and maximums
BbMxH = Betbrain maximum home win odds
BbAvH = Betbrain average home win odds
BbMxD = Betbrain maximum draw odds
BbAvD = Betbrain average draw win odds
BbMxA = Betbrain maximum away win odds
BbAvA = Betbrain average away win odds



Key to total goals betting odds:

BbOU = Number of BetBrain bookmakers used to calculate over/under 2.5 goals (total goals) averages and maximums
BbMx>2.5 = Betbrain maximum over 2.5 goals
BbAv>2.5 = Betbrain average over 2.5 goals
BbMx<2.5 = Betbrain maximum under 2.5 goals
BbAv<2.5 = Betbrain average under 2.5 goals

GB>2.5 = Gamebookers over 2.5 goals
GB<2.5 = Gamebookers under 2.5 goals
B365>2.5 = Bet365 over 2.5 goals
B365<2.5 = Bet365 under 2.5 goals


Key to Asian handicap betting odds:

BbAH = Number of BetBrain bookmakers used to Asian handicap averages and maximums
BbAHh = Betbrain size of handicap (home team)
BbMxAHH = Betbrain maximum Asian handicap home team odds
BbAvAHH = Betbrain average Asian handicap home team odds
BbMxAHA = Betbrain maximum Asian handicap away team odds
BbAvAHA = Betbrain average Asian handicap away team odds

GBAHH = Gamebookers Asian handicap home team odds
GBAHA = Gamebookers Asian handicap away team odds
GBAH = Gamebookers size of handicap (home team)
LBAHH = Ladbrokes Asian handicap home team odds
LBAHA = Ladbrokes Asian handicap away team odds
LBAH = Ladbrokes size of handicap (home team)
B365AHH = Bet365 Asian handicap home team odds
B365AHA = Bet365 Asian handicap away team odds
B365AH = Bet365 size of handicap (home team)

Football-Data would like to acknowledge the following sources which have been utilised in the compilation of Football-Data's results and odds files.

Historical results:
International Soccer Server - http://sunsite.tut.fi/rec/riku/soccer.html
European Football - http://www.eurofootball.be/
RSSSF Archive - http://www.rsssf.com/

Current results (full time, half time)
TBWSport - http://www.tbwsport.com
Livescore- http://www.livescore.com

Match statistics
Sportinglife, ESPN Soccer, Bundesliga.de, Gazzetta.it and Football.fr

Bookmakers betting odds
Betbrain - http://www.betbrain.com
Betbase - http://www.betbase.info

Betting odds for weekend games are collected Friday afternoons, and on Tuesday afternoons for midweek games.

Additional match statistics (corners, shots, bookings, referee etc.) for the 2000/01 and 2001/02 seasons for the English, Scottish and German leagues were provided by Sports.com (now under new ownership and no longer available).

### What does a case represent in each data source, and how many total cases are available?
A case represents one match played and there are 380 total cases in each dataset.

### What are some of the variables that you plan to use?
I plan on using FTAG,and FTHG which are: Full Time Away Goals and Full Time Home goals respectively. I will mutate a new variable FTTG: Full Time Total Goals, which will be the sum of the two variables for each case. I will also create HTTG: Half Time Total Goals to see total number of goals scored at the end of half time. I will also create GD: Goal Differential which will be the difference in score at the end of the game, a negative GD means a loss, and a positive GD means a win, while a 0 GD means a tie. I also renamed all of the non-betting variables I will be including to longer more explanatory variable names. From these variables I will group_by HomeTeam and select variables such that only home team data will be taken, and calculate an average total goals scored for home games of each teach and average goal differential for home games of each team.
```{r}
SerieAClean <- Serie_A_Data %>%
  select(Date,HomeTeam,AwayTeam,FTHG,FTAG,FTR,HTHG,HTAG,HTR,HS,AS,HST,AST,HF,AF,HC,AC,HY,AY,HR,AR) %>%
  mutate(FTTG = FTAG + FTHG,
         HTTG = HTAG + HTHG,
         GD = FTAG - FTHG) %>%
  rename(FullTimeHomeGoals = FTHG, 
         FullTimeAwayGoals = FTAG,
         FullTimeTotalGoals = FTTG,
         FullTimeResult = FTR,
         HalfTimeHomeGoals = HTHG,
         HalfTimeAwayGoals = HTAG,
         HalfTimeTotalGoals = HTTG,
         HalfTimeResult = HTR,
         HomeTeamShots = HS,
         AwayTeamShots = AS,
         HomeTeamShotsOnTarget = HST,
         AwayTeamShotsOnTarget = AST,
         HomeTeamFouls = HF,
         AwayTeamFouls = AF,
         HomeTeamCorners = HC,
         AwayTeamCorners = AC,
         HomeTeamYellowCards = HY,
         AwayTeamYellowCards = AY,
         HomeTeamRedCards = HR,
         AwayTeamRedCards = AR,
         GoalDifferential = GD) %>%
  mutate(matchnum = row_number())
SerieAClean
PremierLeagueClean <- Premier_League_Data %>%
  select(Date,HomeTeam,AwayTeam,FTHG,FTAG,FTR,HTHG,HTAG,HTR,HS,AS,HST,AST,HF,AF,HC,AC,HY,AY,HR,AR) %>%
  mutate(FTTG = FTAG + FTHG,
         HTTG = HTAG + HTHG,
         GD = FTAG - FTHG) %>%
  rename(FullTimeHomeGoals = FTHG, 
         FullTimeAwayGoals = FTAG,
         FullTimeTotalGoals = FTTG,
         FullTimeResult = FTR,
         HalfTimeHomeGoals = HTHG,
         HalfTimeAwayGoals = HTAG,
         HalfTimeTotalGoals = HTTG,
         HalfTimeResult = HTR,
         HomeTeamShots = HS,
         AwayTeamShots = AS,
         HomeTeamShotsOnTarget = HST,
         AwayTeamShotsOnTarget = AST,
         HomeTeamFouls = HF,
         AwayTeamFouls = AF,
         HomeTeamCorners = HC,
         AwayTeamCorners = AC,
         HomeTeamYellowCards = HY,
         AwayTeamYellowCards = AY,
         HomeTeamRedCards = HR,
         AwayTeamRedCards = AR,
         GoalDifferential = GD) %>%
  mutate(matchnum = row_number())
PremierLeagueClean
SerieAHome <- SerieAClean %>%
  select(Date,HomeTeam,FullTimeHomeGoals,FullTimeTotalGoals,FullTimeResult,HalfTimeHomeGoals,HalfTimeTotalGoals,GoalDifferential)%>%
  group_by(HomeTeam) %>%
  summarise(avgTotalGoals = mean(FullTimeTotalGoals),
            avgGoalDiff = mean(GoalDifferential)) %>%
  mutate(matchnum = row_number())
SerieAHome

PremierLeagueHome <- PremierLeagueClean %>%
  select(Date,HomeTeam,FullTimeHomeGoals,FullTimeTotalGoals,FullTimeResult,HalfTimeHomeGoals,HalfTimeTotalGoals,GoalDifferential)%>%
  group_by(HomeTeam) %>%
  summarise(avgTotalGoals = mean(FullTimeTotalGoals),
            avgGoalDiff = mean(GoalDifferential)) %>%
  mutate(matchnum = row_number())
PremierLeagueHome
```



```{r}
ggplot(data=SerieAHome,aes(x=avgTotalGoals))+
  geom_density()

ggplot(data=PremierLeagueHome,aes(x=avgTotalGoals))+
  geom_density()
```
* These two Density plots show that the most common total goals for SerieA games is about 2.45, and most  ommon total goals for Premier League games is around 2.95, so initially it seems that Premier League matches have more goals.


```{r}
ggplot(data=SerieAHome,aes(y=avgTotalGoals))+
  geom_boxplot()

ggplot(data=PremierLeagueHome,aes(y=avgTotalGoals))+
  geom_boxplot()

```
* this Boxplot shows that SerieA games have a wider range with lower minimums for goals scored than games in the Premier League and similar maximums



```{r}
combineddata <- rbind(SerieAHome, PremierLeagueHome)
```



